---
title: 3. Cross-Validation for Quantile Lasso
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Cross-Validation for Quantile Lasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=5)
```

$$
\newcommand{\argmin}{\mathop{\mathrm{argmin}}}
\newcommand{\argmax}{\mathop{\mathrm{argmax}}}
\newcommand{\minimize}{\mathop{\mathrm{minimize}}}
\newcommand{\st}{\mathop{\mathrm{subject\,\,to}}}
$$

Problem setup
===

Consider the problem 
$$
\minimize_{\beta_0,\beta} \; \sum_{i=1}^n \psi_\tau(y_i - \beta_0 - x_i^T \beta) + \lambda \|\beta\|_1
$$
where 
$$
\psi_\tau(v) = \max\{\tau v, (\tau-1) v)\},
$$
often called the "pinball" or "tilted $\ell_1$" loss, for a quantile level $\tau \in (0,1)$. With multiple quantile levels $\tau_k$, $k=1,\ldots,r$, we can simply solve the quantile lasso problem separately for each $\tau_k$.

Cross-validation
===

Suppose we have multiple quantile levels $\tau_k$, $k=1,\ldots,r$ of interest, and we allow  one tuning parameter per quantile level $\lambda_k$, $k=1,\ldots,r$. We seek to minimize the cross-validation (CV) error over the choices of tuning parameters,
$$
\sum_{k=1}^r \sum_{i=1}^n \psi_\tau\Big(y_i - \hat\beta_0(D_{-i}; \tau_k, \lambda_k) - x_i^T \hat\beta(D_{-i}; \tau_k, \lambda_k)\Big),
$$
where $\hat\beta_0(D; \tau, \lambda)$, $\hat\beta(D; \tau, \lambda)$ denotes the quantile lasso solution fit on a data set $D$, at quantile level $\tau$, and tuning parameter value $\lambda$. Above, for each $i=1,\ldots,n$, we use $D_{-i}$ to denote the CV training fold used for point $i$. We can do this just by separately optimizing, for each $k=1,\ldots,r$, the CV error 
$$
\sum_{i=1}^n \psi_\tau\Big(y_i - \hat\beta_0(D_{-i}; \tau_k, \lambda_k) - x_i^T \hat\beta(D_{-i}; \tau_k, \lambda_k)\Big),
$$
over $\lambda_k$. That is, this flexibility---allowing each quantile level its own tuning parameter value---is both statistically and computationlly favorable.

Gaussian example
===

We show a simple example of CV with Gaussian regression data. We also show how to extrapolate to a new set of quantiles, and how to refit at a new set of quantiles.

```{r}
library(quantgen)

set.seed(33)
n = 500
p = 50
x = matrix(rnorm(n*p), n, p)
mu = function(x) x[1] + x[2]
y = apply(x, 1, mu) + rnorm(n)

# Run CV, over just a few quantile levels 
tau = c(0.1, 0.3, 0.5, 0.7, 0.9)
cv_obj = cv_quantile_lasso(x, y, tau=tau, nlambda=30, nfolds=5, lp_solver="gurobi", verbose=TRUE, sort=TRUE)
plot(cv_obj)

# Refit at new quantile levels
tau_new = c(0.01, 0.025, seq(0.05, 0.95, by=0.05), 0.975, 0.99)
new_obj = refit_quantile_lasso(cv_obj, x, y, tau_new, lp_solver="gurobi", verbose=TRUE)

# Predicted and extrapolated quantiles at a few values of x
par(mfrow=c(1,3))
for (i in 1:9) {
  x0 = matrix(rnorm(p), nrow=1)
  qtrue = qnorm(tau_new, mu(x0))
  qpred1 = predict(cv_obj, x0, sort=TRUE)
  qextr1 = quantile_extrapolate(tau, qpred1, tau_new, qfun_left=qnorm, qfun_right=qnorm)
  qpred2 = predict(new_obj, x0, sort=TRUE)
  plot(tau_new, qtrue, type="o", ylim=range(qtrue, qextr1, qpred2, na.rm=TRUE), ylab="Quantile")
  lines(tau_new, qextr1, col=2, pch=20, type="o")
  points(tau, qpred1, col=4, cex=1.5, lwd=2)
  lines(tau_new, qpred2, col=3, pch=20, type="o")
  legend("topleft", legend=c("True", "Predicted", "Extrapolated", "Predicted (refit)"), 
         col=c(1,4,2,3), pch=c(21,21,20,20))
}
```

Poisson example
===

We show another simple example of CV now with Poisson regression data. Through this, we can also demonstrate how to use built-in transform (and inverse transform) functionality. And again we show how to extrapolate to a new set of quantiles, and how to refit at a new set of quantiles.

```{r}
n = 500
p = 50
x = matrix(rnorm(n*p), n, p)
mu = function(x) x[1] + x[2]
y = rpois(n, exp(apply(x, 1, mu)))

# Run CV, over just a few quantile levels 
tau = c(0.1, 0.3, 0.5, 0.7, 0.9)
cv_obj1 = cv_quantile_lasso(x, y, tau=tau, nlambda=30, nfolds=5, lp_solver="gurobi", verbose=TRUE, sort=TRUE)
cv_obj2 = cv_quantile_lasso(x, y, tau=tau, nlambda=30, nfolds=5, lp_solver="gurobi", verbose=TRUE, sort=TRUE,
                            transform=log_pad(a=1), inv_trans=inv_log_pad(a=1)) 
plot(cv_obj1)
plot(cv_obj2)

# Refit at new quantile levels
tau_new = c(0.01, 0.025, seq(0.05, 0.95, by=0.05), 0.975, 0.99)
new_obj1 = refit_quantile_lasso(cv_obj1, x, y, tau_new, lp_solver="gurobi", verbose=TRUE)
new_obj2 = refit_quantile_lasso(cv_obj2, x, y, tau_new, lp_solver="gurobi", verbose=TRUE)

# Predicted and extrapolated quantiles at a few values of x
par(mfrow=c(1,3))
for (i in 1:9) {
  x0 = matrix(rnorm(p), nrow=1)
  qtrue = qpois(tau_new, exp(mu(x0)))
  qpred1 = predict(cv_obj1, x0, sort=TRUE, nonneg=TRUE, round=TRUE)
  qextr1 = quantile_extrapolate(tau, qpred1, tau_new, qfun_left=qpois, qfun_right=qpois, nonneg=TRUE, round=TRUE)
  qpred2 = predict(cv_obj2, x0, sort=TRUE, nonneg=TRUE, round=TRUE)
  qextr2 = quantile_extrapolate(tau, qpred2, tau_new, qfun_left=qpois, qfun_right=qpois, nonneg=TRUE, round=TRUE)
  plot(tau_new, qtrue, type="o", ylim=range(qtrue, qextr1, qextr2, na.rm=TRUE), ylab="Quantile")
  lines(tau_new, qextr1, col=2, pch=20, type="o")
  points(tau, qpred1, col=4, cex=1.5, lwd=2)
  lines(tau_new, qextr2, col=3, pch=20, type="o")
  points(tau, qpred2, col=5, cex=1.5, lwd=2)
  legend("topleft", legend=c("True", "Predicted", "Extrapolated", "Predicted (log)", "Extrapolated (log)"), 
         col=c(1,4,2,5,3), pch=c(21,21,20,21,20))
}

# Refitted versions 
par(mfrow=c(1,3))
for (i in 1:9) {
  x0 = matrix(rnorm(p), nrow=1)
  qtrue = qpois(tau_new, exp(mu(x0)))
  qpred3 = predict(new_obj1, x0, sort=TRUE, nonneg=TRUE, round=TRUE)
  qpred4 = predict(new_obj2, x0, sort=TRUE, nonneg=TRUE, round=TRUE)
  plot(tau_new, qtrue, type="o", ylim=range(qtrue, qpred3, qpred4, na.rm=TRUE), ylab="Quantile")
  lines(tau_new, qpred3, col=3, pch=20, type="o")
  lines(tau_new, qpred4, col=6, pch=20, type="o")
  legend("topleft", legend=c("True", "Predicted (refit)", "Predicted (log, refit)"), col=c(1,3,6), pch=c(21,20,20))
}
```

Noncrossing constraints
===

We refit the last CV object at new quantile levels, and additionally enforce noncrossing constraints. (Note that these noncrossing constraints can only be applied when the values in the `tau` vector being passed to `quantile_genlasso()` are distinct, and sorted in increasing order.)

```{r}
n0 = 9
x0 = matrix(rnorm(n0*p), n0, p)
new_obj3 = refit_quantile_lasso(cv_obj2, x, y, tau_new, noncross=TRUE, x0=rbind(x,x0), lp_solver="gurobi", verbose=TRUE)
qpred1 = predict(new_obj2, x0, sort=FALSE, nonneg=TRUE, round=TRUE)
qpred2 = predict(new_obj2, x0, sort=TRUE, nonneg=TRUE, round=TRUE) 
qpred3 = predict(new_obj3, x0, sort=FALSE, nonneg=TRUE, round=TRUE) # Should be sorted already

par(mfrow=c(1,3))
for (i in 1:9) {
  qtrue = qpois(tau_new, exp(mu(x0[i,])))
  plot(tau_new, qtrue, type="o", ylim=range(qtrue, qpred1[i,], qpred2[i,], qpred3[i,], na.rm=TRUE), ylab="Quantile")
  lines(tau_new, qpred1[i,], col=2, pch=20, type="o")
  lines(tau_new, qpred2[i,], col=3, pch=20, type="o")
  lines(tau_new, qpred3[i,], col=4, pch=20, type="o")
  legend("topleft", legend=c("True", "Predicted", "Predicted (sort)", "Predicted (noncross)"), col=1:4, pch=c(21,20,20,20))
}
```