---
title: 2. Simple Tests for Quantile Lasso
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Simple Tests for Quantile Lasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=5)
```

$$
\newcommand{\argmin}{\mathop{\mathrm{argmin}}}
\newcommand{\argmax}{\mathop{\mathrm{argmax}}}
\newcommand{\minimize}{\mathop{\mathrm{minimize}}}
\newcommand{\st}{\mathop{\mathrm{subject\,\,to}}}
$$

Problem setup
===

Consider the problem 
$$
\minimize_{\beta_0,\beta} \; \sum_{i=1}^n \psi_\tau(y_i - \beta_0 - x_i^T \beta) + \lambda \|\beta\|_1
$$
where 
$$
\psi_\tau(v) = \max\{\tau v, (\tau-1) v)\},
$$
often called the "pinball" or "tilted $\ell_1$" loss, for a quantile level $\tau \in (0,1)$. We can rewrite this succintly letting $y$ be the response vector with elements $y_i$, and $X$ the predictor matrix with rows $x_i$, as
$$
\minimize_\beta \; \psi_\tau(y-X\beta) + \lambda \|\beta\|_1.
$$
Here we interpret $\psi_\tau$ as being applied componentwise to $y-X\beta$, and we suppress the notation for the intercept; as usual, to include an intercept, we just append a columns of 1s to $X$.

LP reformulation
===

We can reformulate this as an LP:
\begin{alignat*}{2}
&\minimize_{\beta,u,v} \quad && 1_n^\top v + \lambda 1_p^\top u\\
&\st \quad && v \geq \tau (y-X\beta), \\
&&& v \geq (\tau-1) (y-X\beta), \\
&&& u \geq \beta, \; u \geq -\beta.
\end{alignat*}
Here $1_k$ is the all 1s vector of dimension $k$, and all inequalities are interpreted componentwise. 

Simple tests
===

We run a bunch of simple tests against `rqPen`.

```{r}
library(rqPen)
library(quantgen)

set.seed(33)
n = 500
p = 50
x = matrix(rnorm(n*p), n, p)
x0 = matrix(rnorm(n*p), n,)
mu = function(x) x[1] + x[2]
y = apply(x, 1, mu) + rnorm(n)

tau = 0.8
lambda = sqrt(get_lambda_max(x, y, Matrix::Diagonal(p)))

# No intercept, no standardize
a1 = quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=FALSE, lp_solver="gurobi")
a2 = quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=FALSE, lp_solver="glpk")
a3 = rq.lasso.fit(x, y, tau, lambda/n, intercept=FALSE, scalex=FALSE)
max(abs(coef(a1)-coef(a3)))
max(abs(coef(a2)-coef(a3)))
max(abs(predict(a1,x0) - predict(a3,x0)))
max(abs(predict(a2,x0) - predict(a3,x0)))

# Yes intercept, no standardize
a1 = quantile_lasso(x, y, tau, lambda, intercept=TRUE, standardize=FALSE, lp_solver="gurobi")
a2 = quantile_lasso(x, y, tau, lambda, intercept=TRUE, standardize=FALSE, lp_solver="glpk")
a3 = rq.lasso.fit(x, y, tau, lambda/n, intercept=TRUE, scalex=FALSE)
max(abs(coef(a1)-coef(a3)))
max(abs(coef(a2)-coef(a3)))
max(abs(predict(a1,x0) - predict(a3,x0)))
max(abs(predict(a2,x0) - predict(a3,x0)))

# No intercept, yes standardize
a1 = quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=TRUE, lp_solver="gurobi")
a2 = quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=TRUE, lp_solver="glpk")
a3 = rq.lasso.fit(x, y, tau, lambda/n, intercept=FALSE, scalex=TRUE)
max(abs(coef(a1)[-1]-coef(a3)))
max(abs(coef(a2)[-1]-coef(a3)))
max(abs(predict(a1,x0) - predict(a3,x0)))
max(abs(predict(a2,x0) - predict(a3,x0)))
# I think that rqPen is forgetting to add the result of centering the x's back into the intercept

# Yes intercept, yes standardize
a1 = quantile_lasso(x, y, tau, lambda, intercept=TRUE, standardize=TRUE, lp_solver="gurobi")
a2 = quantile_lasso(x, y, tau, lambda, intercept=TRUE, standardize=TRUE, lp_solver="glpk")
a3 = rq.lasso.fit(x, y, tau, lambda/n, intercept=TRUE, scalex=TRUE)
max(abs(coef(a1)[-1]-coef(a3)[-1]))
max(abs(coef(a2)[-1]-coef(a3)[-1]))
max(abs(predict(a1,x0) - predict(a3,x0)))
max(abs(predict(a2,x0) - predict(a3,x0)))
# I think that rqPen is forgetting to add the result of centering the x's back into the intercept
```

Speed comparison
===

On a bigger problem, we compare speeds across LP solvers (Gurobi and GLPK), and also `rqPen`.

```{r}
n = 2000
p = 1000
x = matrix(rnorm(n*p), n, p)
mu = function(x) x[1] + x[2]
y = apply(x, 1, mu) + rnorm(n)

tau = 0.8
lambda = sqrt(get_lambda_max(x, y, Matrix::Diagonal(p)))

system.time(quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=FALSE, lp_solver="gurobi", verbose=TRUE))[3]
system.time(quantile_lasso(x, y, tau, lambda, intercept=FALSE, standardize=FALSE, lp_solver="glpk", verbose=TRUE))[3]
system.time(rq.lasso.fit(x, y, tau, lambda/n, intercept=FALSE, scalex=FALSE))[3]
```