---
title: Getting started with quantgen
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with quantgen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

$$
\newcommand{\argmin}{\mathop{\mathrm{argmin}}}
\newcommand{\argmax}{\mathop{\mathrm{argmax}}}
\newcommand{\minimize}{\mathop{\mathrm{minimize}}}
\newcommand{\st}{\mathop{\mathrm{subject\,\,to}}}
$$

This package provides tools for generalized quantile modeling: regularized 
quantile regression (with generalized lasso penalties and noncrossing 
constraints), cross-validation, quantile extrapolation, and quantile ensembles. 

## Installing

This package is not on CRAN yet, so it can be installed using the
[`devtools`](https://cran.r-project.org/package=devtools) package:

```{r, eval = FALSE}
devtools::install_github(repo="ryantibs/quantgen", subdir="R-package/quantgen")
```

Building the vignettes takes a substantial amount of time. They are not included
in the package by default. If you want to include vignettes (for local access), 
then you can use:

```{r, eval = FALSE}
devtools::install_github(repo="ryantibs/quantgen", subdir="R-package/quantgen",
                         build_vignettes=TRUE, dependencies=TRUE)
```

## Quantile lasso

Here we give some basic examples of how to fit a quantile lasso model. For
details on how this problem is defined (and how `quantgen` casts it into a 
linear program), see the [mathematical details vignette]. 

To fit a quantile lasso model, we can use `quantile_lasso()`. Below, we do so
at the quantile level $\tau = 0.8$: 

```{r}
library(quantgen)

set.seed(0)
n = 100
p = 10
x = matrix(rnorm(n*p), n, p)
x0 = rnorm(p)
mu = function(x) 2 + x[1] + x[2]
y = apply(x, 1, mu) + rnorm(n)

tau = 0.8
lambda = 2 * sqrt(get_lambda_max(x, y, Matrix::Diagonal(p)))
obj = quantile_lasso(x, y, tau=tau, lambda=lambda)

class(obj)
round(coef(obj), 3) # Vector of length p + 1 (first entry is the intercept)
predict(obj, newx=x0)
```

The default in `quantile_lasso()` is to include an intercept in the model (via
`intercept = TRUE`) and to standardize the predictors (to have zero mean and 
unit norm, via `standardize = TRUE`). **Important note** (obvious in hindsight, 
but important nonetheless): in a quantile model, the intercept cannot be omitted
by simply centering the response and the predictors, as it can in a standard 
regression model. The intercept depends critically on the quantile level $\tau$.
Above, the fact that the estimated intercept exceeds 2 is *not* an instance of 
poor estimation in the quantile lasso model, but reflects the gap between the 
level 0.8 and 0.5 quantiles of the standard normal distribution (note that 
`qnorm(0.8)` $\approx 0.84$).

The `get_lambda_max()` function computes (approximately) the effective maximum 
of the regularization path for a penalized quantile model with $\tau = 0.5$,
given a predictor matrix, response vector, and penalty matrix. In the case of 
the lasso, the penalty matrix is the identity.

We can see that the `quantile_lasso()` function returns an object of class 
`quantile_lasso` (inherited from `quantile_genlasso`), and comes with associated
utilities `coef()` and `predict()`.

## Multiple quantiles

To fit solutions at multiple quantile levels, we can simply pass a vector for 
the `tau` argument:

```{r}
tau_vec = c(0.1, 0.5, 0.9)
obj = quantile_lasso(x, y, tau=tau_vec, lambda=lambda)

round(coef(obj), 3) # Matrix of dimension (p + 1) x (number of tau values)
predict(obj, newx=x0)
```

To fit solutions at multiple tuning parameter values, we can also pass a vector 
for the `lambda` argument, in which case `quantile_lasso()` internally recycles 
the `tau` and `lambda` arguments so that they have equal length, and then solves 
separate quantile lasso problems, each with a quantile level `tau[i]` and tuning
parameter `lambda[i]`. The two most common use cases are to pass a single value
for `tau` and a vector for `lambda`, or to pass a vector for `tau` and a single
value for `lambda`. 

## Tau x lambda grid

Solving quantile lasso problems over a two-dimensional grid of quantile levels
and tuning parameter values is most convenient with `quantile_lasso_grid()`. 
In this function, passing `lambda` is optional; when not specified, the function 
`quantile_lasso_grid()` internally sets `lambda` to be a vector of length
`nlambda = 30`, log-spaced between the value computed by `get_lambda_max()`
and `lambda_min_ratio = 0.001` times this value. 

```{r}
obj = quantile_lasso_grid(x, y, tau_vec)

# Matrix of dimension (p + 1) x (number of tau and lambda pairs)
dim(coef(obj)) 

# Array of dimension (number of x0 points) x (number of lambda values) x 
# (number of tau values)
dim(predict(obj, new=x0)) 
```

## Other options

The `quantile_lasso()` function is a special case of its a parent function
`quantile_genlasso()`, which allows for specification of a general penalty
matrix. Both `quantile_lasso()` and `quantile_genlasso()` have several other
options that you can read about in their documentation: 

- observation weights, which can be set via `weights`;
- noncrossing constraints, which can be set via `noncross` and `x0`;
- faster optimization using Gurobi (requires separate installation of the 
  `gurobi` package), which can be set with `lp_solver`;
- response transformation pre-optimization (and inverse transformation 
  post-prediction), which can be set with `transform` (and `inv_trans`).